Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-02-18 vertical-thermal-review.Rmd
vertical lines and grey shading should be removed from Fig.4
#107
proofreading: Global change responses section

![**Figure `r fig_treerings`. Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species. Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET) for the maximum and minimum tails of the DBH distribution included in the analysis.  Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction. Other model terms are held constant at their mean. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable. From @anderson-teixeiraJointEffectsClimate2022. In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests. Growth is expressed as relativized basal area increment (BAI), where values > and < 100% indicate higher- or lower- than-average growth, respectively. Again, colored lines indicate modeled mean temperature responses and transparent ribbons indicate 95% confidence intervals. From @rollinsonClimateSensitivityUnderstory2020.](tree_rings/Figure_4_tree_rings.png)

@@

To the extent that warming is coupled to drought, we expect that the largest trees will usually be hardest-hit, particularly in cases where the drought is severe, and that this will occur at least in part because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs. `r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimate2022].
Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and accompanying VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.
In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015;@hoganDroughtInterannualVariability2019].
An interesting open question is to what extent these patterns vary with the nature of the drought, including the relative contributions to stress from low soil moisture versus high VPD. 
While the two are coupled over longer time scales [@humphreySoilMoistureAtmosphere2021; @novickIncreasingImportanceAtmospheric2016], the latter can be intense for short periods even when soil moisture is high (e.g., during a heat wave) and exerts a stronger influence on ET in many biomes [@novickIncreasingImportanceAtmospheric2016]. 
Given the mechanisms reviewed here, we might expect that atmospheric dryness in particular skews the stress more towards the exposed canopy trees.

What remains poorly understood is how responses to warming will vary across vertical gradients under mesic conditions. 
The leaf-level responses are relatively well-understood: at the metabolic level, photosynthesis and respiration are coupled to photosynthetic thermal optima, which reflect adaptation to the local climate but do not vary consistently with height in the vertical profile (see section 4.2, Table 2). 
Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021]. 
While leaves display substantial plasticity to adapt to warmer temperatures [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], failure to fully acclimate will result in reduced carbon sequestration at leaf and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. 
Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019]. 
Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach thermal tolerance thresholds during heat waves, resulting in photosynthetic decline or even leaf death [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress].

@@

As discussed above, rising temperatures and severe droughts place canopy trees at particularly elevated risk of mortality. 
Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting,  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change[@ipccClimateChange20212021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011]. 
Increase in the severity and frequency of heat waves, accompanied with an increase in VPD and ET, can exacerbate effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].
Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases incident radiation levels and temperatures within the forest [@juckerCanopyStructureTopography2018; @starkReframingTropicalSavannization2020]. 
Such changes often result in increased growth of smaller trees, which benefit from increased light [@hoganDroughtInterannualVariability2019;@bennettLargerTreesSuffer2015], and wetter forests can prove quite resilient to canopy disturbance [@millerReducedImpactLogging2011].
However, this shift to hotter and drier microclimates makes some forests more susceptible to further disturbances, for example, increasing fire risk [@brandoAbruptIncreasesAmazonian2014; @aragao21stCenturyDroughtrelated2018]. 
Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but may have reduced probability of recovering and persisting when affected by severe canopy disturbance [@tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @mcdowellPervasiveShiftsForest2020; @floresFloodplainsAchillesHeel2017].

>>>>

![**Figure `r fig_treerings`. Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species. Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET) for the maximum and minimum tails of the DBH distribution included in the analysis.  Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction. Other model terms are held constant at their mean. Transparent ribbons indicate 95% confidence intervals. From @anderson-teixeiraJointEffectsClimate2022. In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests. Growth is expressed as relativized basal area increment (BAI), where values > and < 100% indicate higher- or lower- than-average growth, respectively. Again, colored lines indicate modeled mean temperature responses and transparent ribbons indicate 95% confidence intervals. From @rollinsonClimateSensitivityUnderstory2020.](tree_rings/Figure_4_tree_rings.png)

@@

To the extent that warming is coupled to drought, we expect that the tallest trees will usually be hardest-hit, particularly in cases where the drought is severe, and that this will occur at least in part because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs. `r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimate2022].
Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and accompanying VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.
In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015;@hoganDroughtInterannualVariability2019].
An interesting open question is to what extent these patterns vary with the nature of the drought, including the relative contributions to stress from low soil moisture versus high VPD. 
While the two are coupled over longer time scales [@humphreySoilMoistureAtmosphere2021; @novickIncreasingImportanceAtmospheric2016], the latter can be intense for short periods even when soil moisture is high (e.g., during a heat wave) and exerts a stronger influence on ET in many biomes [@novickIncreasingImportanceAtmospheric2016]. 
Given the mechanisms reviewed here, we might expect that atmospheric dryness in particular skews the stress more towards the exposed canopy trees.

What remains poorly understood is how responses to warming will vary across vertical gradients under mesic conditions. 
The leaf-level responses are relatively well-understood: at the metabolic level, photosynthesis and respiration are coupled to photosynthetic thermal optima, which reflect adaptation to the local climate but do not vary consistently with height in the vertical profile (see section 4.2, Table 2). 
Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021]. 
While leaves display substantial plasticity to adapt to warmer temperatures [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], failure to fully acclimate will result in reduced carbon sequestration at leaf and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. 
Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019]. 
Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach thermal tolerance thresholds during heat waves, resulting in photosynthetic decline or even leaf death [@osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress].

@@

Rising temperatures and severe droughts place canopy trees at particularly elevated risk of mortality. 
Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting,  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change [@ipccClimateChange20212021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011]. 
Increase in the severity and frequency of heat waves, accompanied with an increase in VPD and ET, can exacerbate effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].
Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases incident radiation levels and temperatures within the forest [@juckerCanopyStructureTopography2018; @starkReframingTropicalSavannization2020]. 
Such changes often result in increased growth of smaller trees, which benefit from increased light [@hoganDroughtInterannualVariability2019;@bennettLargerTreesSuffer2015], and wetter forests can prove quite resilient to canopy disturbance [@millerReducedImpactLogging2011].
However, this shift to hotter and drier microclimates makes some forests more susceptible to further disturbances, for example, increasing fire risk [@brandoAbruptIncreasesAmazonian2014; @aragao21stCenturyDroughtrelated2018]. 
Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but may have reduced probability of recovering and persisting when affected by severe canopy disturbance [@tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @mcdowellPervasiveShiftsForest2020; @floresFloodplainsAchillesHeel2017].
